### 2016-06-27 store-consistency.tex
Example added.

\begin{figure}
\centering
\begin{txnimpcode}
 $\begin{decoration}
 P_1:\{ {\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj\\
        \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2}
        \hboar \C{Wd1} \wedge \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a2} \}
 \end{decoration}$
  txn$\langle$'Wd1'$\rangle${
   $\begin{decoration}
    \phi_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj
      {\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj\\
       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
        {\C{Wd2} \hboar \C{Wd1}} 
       \Rightarrow \C{C = k-a2} \}
    \end{decoration}$ 
    v1 = C
   $\begin{decoration}
    \phi_2 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v1 = k} \conj\\
       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
        {\C{Wd2} \hboar \C{Wd1}} 
       \Rightarrow \C{C = k-a2} \wedge \C{v1 = k-a2}\}
    \end{decoration}$ 
    if (v1 $\ge$ a1) {
      v2 = C;
     $\begin{decoration}
      \phi_3 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v2 = k} \conj\\
       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
          {\C{Wd2} \hboar \C{Wd1}} \Rightarrow \C{C = k-a2} \wedge \\
          \hspace*{1.9in}\C{v2 = k-a2}\}
      \end{decoration}$ 
      v3 = v2 - a1;
     $\begin{decoration}
      \phi_4 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v3 = k-a1} \conj\\
         \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
          {\C{Wd2} \hboar \C{Wd1}} 
         \Rightarrow \C{C = k-a2} \wedge \\
         \hspace*{1.9in}\C{v3 = k-a2-a1}\}
      \end{decoration}$ 
      C := v3
     $\begin{decoration}
      \phi_5 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}) 
                \conj\\
         \hspace*{0.3in}{\committed(\C{Wd2})} 
                \Rightarrow \C{C = k-a1-a2}\}
      \end{decoration}$ 
    }
  }
 $\begin{decoration}
  Q_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}
            \conj \committed(\C{Wd1}) \conj\\
      \hspace*{0.12in}{\committed(\C{Wd2})} 
          \Rightarrow \C{C = k-a1-a2} \}
  \end{decoration}$ 
\end{txnimpcode}

\caption{`Wd1' transaction decorated with assertions}
\label{fig:wd1-decorated}
\end{figure}

The fully decorated implementation of transaction `Wd1' is shown in
Fig.~\ref{fig:wd1-decorated}. Decorated `Wd2' is similar. Assignment
statements are broken down and temporary local variables (\C{v1},
\C{v2} and \C{v3}) are introduced so as to separate shared variable
reads and writes.  Since we reason in terms of executions, all
assertions implicitly refer to the current execution (just as hoare
triples implicitly refer to the current state). Also implicit is the
invariant \C{k $\ge$ a1+a2}. Proposition $\committed(T)$ indicates
that the transaction $T$ is already committed. Precondition ($P_1$) of
`Wd1' accounts for the possibility of `Wd2' committing before `Wd1',
writing \C{k-a2} to \C{C}.  Precondition of `Wd2' is similar.  Since
neither `Wd1' nor `Wd2' are committed at the beginning, the
precondition in Fig.~\ref{fig:motiv-eg-1} is extended to include
$\neg\committed(\C{Wd1}) \wedge \neg\committed(\C{Wd2})$, from which
$P_1$ follows. Once the execution is inside `Wd1', commit of `Wd2'
($\committed(\C{Wd2})$) may mean that either `Wd2' happened before
`Wd1', or that it committed concurrently with `Wd1'. Since SI
proscribes the latter possibility, we only consider the case when
$\C{Wd2} \hboar \C{Wd1}$. A proof for $\C{Wd2} \hboar \C{Wd1}$ is
obtained subsequently, allowing us to get rid of the special case.

The first obligation is to show that the assertions that decorate
`Wd1' are indeed valid. This entails proving that, for each statement
of `Wd1', if evaluating the statement and extending an execution that
satisfies its precondition results in a well-formed execution, then
the extended execution must satisfy the postcondition. The
well-formedness assumption lets us ignore the executions that violate
the trace invariant lead the machine to a stuck state. This is useful
in case of the assignment statement that writes \C{v3} to \C{C}. Since
`Wd2' also writes to \C{C}, the trace invariant ($\I$) requires
$\C{Wd2} \hboar \C{Wd1}$ if $\committed(\C{Wd2})$ (the alternative
$\C{Wd1} \hboar \C{COMMIT}(\C{Wd2})$ allowed by \C{SI(Wd1)} is
impossible in this case because \C{Wd2} has already committed, while
\C{Wd1} is still in progress). This lets us treat the special case
$\committed(\C{Wd2})\conj \C{Wd2} \hboar \C{Wd1}$ as being equivalent
to the general case $\committed(\C{Wd2})$, which is required to prove
the postcondition of `Wd1'. Observe that \C{RC(Wd1)} does not justify
this generalization, hence replacing \C{SI(Wd1)} by \C{RC(Wd1)} is not
sufficient to prove the postcondition of `Wd1'.

The second obligation is to prove that assertions are valid despite
the interference from the the concurrent thread executing `Wd2'. The
interference is given by the following rely relation ($R_1$):
\begin{smathpar}
\begin{array}{lcl}
  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd2})} \conj 
        \underE{\I} \conj \E'\Vdash \I \conj\\
%       \conj (\E'-\E) \subseteq \C{Wd2} \conj \\
%   & & \hspace*{0.5in} \E' \Vdash \committed(\C{Wd2}) ~\Rightarrow~ \E'
%       \Vdash \C{Wd2} \wrstoar \C{C} \conj \\ 
%   & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj \E' \Vdash
%   \committed(\C{Wd2}) \Rightarrow \C{C=k-a1-a2} \\
    & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
        \C{COMMIT(Wd2)} \in (\E'-\E) \\
    & & \hspace*{0.8in}\Rightarrow \C{C=k-a2} \conj 
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \conj \\
    & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj
        \C{COMMIT(Wd2)} \in (\E'-\E) \\
    & & \hspace*{0.6in}\Rightarrow \C{C=k-a1-a2} \conj
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \}\\
\end{array}
\end{smathpar}
The rely relation says the following about the concurrent thread: (1).
It may interfere only if `Wd2' is not already committed, (2). Any
interference takes well-formed executions to well-formed executions,
(3). If the interference commits `Wd2', then `Wd2' should have already
written to \C{C}, and (4). The value written is either $\C{k-a1-a2}$
or $\C{k-a1}$ depending on whether or not `Wd1' has already committed.
Interference from `Wd2' is allowed between any two operations of
`Wd1', and assertions are required to be valid despite the
interference. For instance, if the execution ($\E$) immediately after
evaluating $\C{v1 = C}$ satisfies $\phi_2$ (i.e., $\underE{\phi_2}$),
and an interference from `Wd2' extends $\E$ to $\E'$ (i.e.,
$R_1(\E,\E')$), then the execution $\E'$ just before evaluating the
condition must satisfy $\phi_2$ (i.e., $\E' \Vdash \phi_2$). Indeed,
all $\phi_i$'s do satisfy this \emph{stability} condition, thus
lending the validity to the proof.

>>>>

We shall now consider the proof of the example in
Fig.~\ref{fig:motiv-eg-1} in greater detail. We assume an {\sc sc}
store, such as a relational database, whose store-specific trace
invaraint ($\I_s$) was shown previously. Both transactions are run at
{\sc si} isolation, hence $\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}}
\conj \underE{\C{SI(Wd1)}}$. As usual, $\I$ is $\I_s \conj \I_c$.

\begin{figure}
\centering
\begin{txnimpcode}
 $\begin{decoration}
 P_1:\{ {\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj\\
        \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2}
        \visar \C{Wd1} \wedge \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a2} \}
 \end{decoration}$
  txn$\langle$'Wd1'$\rangle${
   $\begin{decoration}
    \phi_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj
      {\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj\\
       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
        {\C{Wd2} \visar \C{Wd1}} 
       \Rightarrow \C{C = k-a2} \}
    \end{decoration}$ 
    v1 = C
   $\begin{decoration}
    \phi_2 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v1 = k} \conj\\
       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
        {\C{Wd2} \visar \C{Wd1}} 
       \Rightarrow \C{C = k-a2} \wedge \C{v1 = k-a2}\}
    \end{decoration}$ 
    if (v1 $\ge$ a1) {
      v2 = C;
     $\begin{decoration}
      \phi_3 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v2 = k} \conj\\
       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
          {\C{Wd2} \visar \C{Wd1}} \Rightarrow \C{C = k-a2} \wedge \\
          \hspace*{1.9in}\C{v2 = k-a2}\}
      \end{decoration}$ 
      v3 = v2 - a1;
     $\begin{decoration}
      \phi_4 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v3 = k-a1} \conj\\
         \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
          {\C{Wd2} \visar \C{Wd1}} 
         \Rightarrow \C{C = k-a2} \wedge \\
         \hspace*{1.9in}\C{v3 = k-a2-a1}\}
      \end{decoration}$ 
      C := v3
     $\begin{decoration}
      \phi_5 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}) 
                \conj\\
         \hspace*{0.3in}{\committed(\C{Wd2})} 
                \Rightarrow \C{C = k-a1-a2}\}
      \end{decoration}$ 
    }
  }
 $\begin{decoration}
  Q_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}
            \conj \committed(\C{Wd1}) \conj\\
      \hspace*{0.12in}{\committed(\C{Wd2})} 
          \Rightarrow \C{C = k-a1-a2} \}
  \end{decoration}$ 
\end{txnimpcode}

\caption{\C{Wd1} transaction decorated with assertions}
\label{fig:wd1-decorated}
\end{figure}

The fully decorated implementation of transaction \C{Wd1} is shown in
Fig.~\ref{fig:wd1-decorated}. Decorated \C{Wd2} is similar and not
shown. Assignment statements are broken down and temporary local
variables (\C{v1}, \C{v2} and \C{v3}) are introduced so as to separate
shared variable reads and writes. All assertions implicitly refer to
the current execution ($\E$), just as hoare triples implicitly refer
to the current state. The context for propositions is also the
implicit, i.e., we write $\psi$ instead of $\underE{\psi}$. The
proposition \C{k $\ge$ a1+a2} remains an invariant, hence elided. The
precondition ($P_1$) of \C{Wd1} accounts for the possibility of
\C{Wd2} committing before \C{Wd1}, writing \C{k-a2} to \C{C}.
Precondition of \C{Wd2} is similar.  Since neither \C{Wd1} nor \C{Wd2}
are committed at the beginning, the execution-based assertion
corresponding to the precondition ($P$) of Fig.~\ref{fig:motiv-eg-1}
must include $\neg\committed(\C{Wd1}) \wedge \neg\committed(\C{Wd2})$,
from which $P_1$ follows. Once the execution is inside \C{Wd1}, commit
of \C{Wd2} ($\committed(\C{Wd2})$) may mean that either \C{Wd2}
happened before \C{Wd1} (hence can be visible to all of \C{Wd1}), or
that it committed concurrently with \C{Wd1} (hence cannot be visible
to all of \C{Wd2}).  Since SI proscribes the latter possibility, we
only consider the case when $\C{Wd2} \visar \C{Wd1}$. A proof for
$\C{Wd2} \visar \C{Wd1}$ is obtained subsequently, allowing us to get
rid of the special case.

First, we focus on the sequential aspect of the proof and show that
the assertions that decorate \C{Wd1} are indeed valid. The proof for
each triple follows from the rule for the corresponding term in
Fig.~\ref{fig:rg-rules}. For illustration, we consider the triple
corresponding to the assignment statement (line ??). The precondition
($\phi_4$) asserts that \C{Wd2} writes to \C{C}, and the value of the
temporary variable \C{v3} is \C{k-a2-a1}. The RG rule
(\rulelabel{RG-Asgn}) for assignment statement allows us to easiliy
conclude the following about the execution state after the assignment:
\begin{smathpar}
\begin{array}{l}
    {\neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{C} 
    \conj (\neg\committed(\C{Wd2})} \Rightarrow \C{C = k}) \conj\\
    ({\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar
    \C{C}) \conj \\
    ({\committed(\C{Wd2})} \wedge {\C{Wd2} \visar \C{Wd1}} 
    \Rightarrow \C{C = k-a2} \wedge \C{C = k-a2-a1})
\end{array}
\end{smathpar}
The rule also lets us assume that the execution satisfies trace
invariant ($\I$), which asserts {\sc si} for both transactions.  Since
both transactions write to \C{C}, {\sc si} requires either $\C{Wd1}
\visar \C{Wd2}$, or $\C{Wd2} \visar \C{Wd1}$. Since \C{Wd1} is not yet
committed ($\neg\committed(\C{Wd1})$), {\sc si} prohibits the former
possibility, allowing us the deduce the $\C{Wd2} \visar \C{Wd1}$. This
lets us derive ${\committed(\C{Wd2})}  \Rightarrow \C{C = k-a2} \wedge
\C{C = k-a2-a1}$, allowing us to prove the postcondition ($\phi_5$).

The second part of the proof is to show that assertions are stable
despite the interference from the the concurrent thread executing
\C{Wd2}. The interference is given by the following rely relation
($R_1$):
\begin{smathpar}
\begin{array}{lcl}
  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd2})} \conj 
        \underE{\I} \conj \E'\Vdash \I \conj\\
%       \conj (\E'-\E) \subseteq \C{Wd2} \conj \\
%   & & \hspace*{0.5in} \E' \Vdash \committed(\C{Wd2}) ~\Rightarrow~ \E'
%       \Vdash \C{Wd2} \wrstoar \C{C} \conj \\ 
%   & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj \E' \Vdash
%   \committed(\C{Wd2}) \Rightarrow \C{C=k-a1-a2} \\
    & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
        \C{COMMIT(Wd2)} \in (\E'-\E) \\
    & & \hspace*{0.8in}\Rightarrow \C{C=k-a2} \conj 
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \conj \\
    & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj
        \C{COMMIT(Wd2)} \in (\E'-\E) \\
    & & \hspace*{0.6in}\Rightarrow \C{C=k-a1-a2} \conj
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \}\\
\end{array}
\end{smathpar}
The rely relation says the following about the concurrent thread: (1).
It may interfere only if \C{Wd2} is not already committed, (2). Any
interference takes well-formed executions to well-formed executions,
(3). If the interference commits \C{Wd2}, then \C{Wd2} should have already
written to \C{C}, and (4). The value written is either $\C{k-a1-a2}$
or $\C{k-a1}$ depending on whether or not \C{Wd1} has already committed.
Stability for most assertions is straightforward. The only non-trivial
proof is for $\phi_5$, where {\sc si} condition should be used to
show that any interference from \C{Wd2} is invalid. The proof proceeds
on the same lines as the Hoare proof discussed above; it is also
discussed in \S\ref{sec:motivation}.

The final aspect of the proof is to show that any interference from
\C{Wd1} is contained in its guarantee relation $G_1$, which becomes
the rely relation ($R_2$) for \C{Wd2}. $R_2$ (hence $G_1$) is same as
$R_1$ shown above, but with \C{Wd2} and \C{Wd1} interchanged. $\ldots$
